---
title: "Association between COVID-19 and Marathon Records"
author: "BST 260 Team 11"
date: "as of 12/3/2021"
output: html_document
---

## Introduction/Backgrounds

Intro and something here. 

## Data Collection Part (Yusuke Yoshikawa)

```{r, message=FALSE, warning=FALSE}
#### Libraries ####
library(tidyverse)
library(rvest)
library(lubridate)
#if (!require(RCurl)) {install.packages("RCurl", dependencies=TRUE)}
#library(RCurl)
```

```{r, message=FALSE, warning=FALSE}
#### Men at Tokyo 2020 ####
# URL of Men at Tokyo2020
url_20m <- "https://en.wikipedia.org/wiki/Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Men%27s_marathon"

# Extract all tables in the page
tab <- read_html(url_20m) %>% 
  html_nodes("table") 
# Table of interest
dat_20m <- tab[[8]] %>% 
  html_table() 

# Data reshaping
colnames(dat_20m)[1:4] <- c("rank", "athlete", "country", "time")
dat_20m$rank[1:3] <- c(1:3)
dat_20m <- dat_20m %>% 
  mutate(rank = ifelse(rank=="—", NA, rank)) %>% 
  mutate(sb = ifelse(is.na(rank), NA, 0)) %>% 
  mutate(sb = ifelse(Notes %in% c("SB","NR","PB"), 1, sb)) %>% 
  mutate(dnf = ifelse(Notes %in% c("DNF","DNS","DSQ"), 1, 0)) %>% 
  mutate(time = lubridate::hms(time)) %>% 
  select(rank, athlete, country, time, sb, dnf)

# Age data collection
urls <- read_html(url_20m) %>% 
  html_nodes("table") %>% 
  .[[8]] %>% 
  html_nodes("a[href *= '/wiki/' ]") %>% # extract the individual athletes' wikis
  html_attr("href") %>% 
  .[!str_detect(.,"Olympics") & !str_detect(., "Bests") & !str_detect(., "conditions") & !str_detect(., "records")] %>%
  paste("https://en.wikipedia.org", ., sep="")

# Extraction of Wiki text function
text_detect <- function(url){
  read_html(url) %>% 
    html_nodes("body") %>% 
    .[[1]] %>% 
    html_text()
}
wiki_text <- sapply(urls, text_detect)

# DOB detection by string processing
dob_detect <- function(string){
  string %>% str_extract(
  "(born \\d{1,2} (January|February|March|April|May|June|July|August|September|October|November|December).\\d{4})|(born (January|February|March|April|May|June|July|August|September|October|November|December) \\d{1,2}. \\d{4})"
  )
}
dobs <- sapply(wiki_text, dob_detect) %>% 
  as.character() %>% 
  str_replace("born ", "")
dobs_d <- dobs %>% str_extract("\\d{1,2}")
dobs_m <- dobs %>% str_extract("(January|February|March|April|May|June|July|August|September|October|November|December)")
dobs_y <- dobs %>% str_extract("\\d{4}")
dobs <- paste(dobs_y, dobs_m, dobs_d, sep="-") %>% ymd()
dat_20m$dob <- dobs

# Calculate age at Tokyo 2020
if (!require(eeptools)) {install.packages("eeptools", dependencies=TRUE)}
dat_20m <- dat_20m %>% 
  mutate(age = eeptools::age_calc(dob, ymd("2021-08-08"), units = "years") %>% floor)

# Sex category 
dat_20m$sex <- "Men"
# Game label
dat_20m$olympic <- "Tokyo2020"

```

```{r, message=FALSE, warning=FALSE}
#### Women at Tokyo 2020 ####
# URL of Women at Tokyo 2020
url_20w <- "https://en.wikipedia.org/wiki/Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Women%27s_marathon"

# Extract all tables in the page
tab <- read_html(url_20w) %>% 
  html_nodes("table") 
# Table of interest
dat_20w <- tab[[8]] %>% 
  html_table()

# Data reshaping
colnames(dat_20w)[1:4] <- c("rank", "athlete", "country", "time")
dat_20w$rank[1:3] <- c(1:3)
dat_20w <- dat_20w %>% 
  mutate(rank = ifelse(rank=="–", NA, rank)) %>% 
  mutate(sb = ifelse(is.na(rank), NA, 0)) %>% 
  mutate(sb = ifelse(Notes %in% c("SB","NR","PB"), 1, sb)) %>% 
  mutate(dnf = ifelse(Notes %in% c("DNF","DNS","DSQ"), 1, 0)) %>% 
  mutate(time = lubridate::hms(time)) %>% 
  select(rank, athlete, country, time, sb, dnf)

# Age data collection
urls <- read_html(url_20w) %>% 
  html_nodes("table") %>% 
  .[[8]] %>% 
  html_nodes("a[href *= '/wiki/' ]") %>% # extract the individual athletes' wikis
  html_attr("href") %>% 
  .[!str_detect(.,"Olympics") & !str_detect(., "Bests") & !str_detect(., "conditions") & !str_detect(., "records")] %>%
  paste("https://en.wikipedia.org", ., sep="")

# Extraction of Wiki text by the function above
wiki_text <- sapply(urls, text_detect)

# DOB detection by string processing
dobs <- sapply(wiki_text, dob_detect) %>% 
  as.character() %>% 
  str_replace("born ", "")
dobs_d <- dobs %>% str_extract("\\d{1,2}")
dobs_m <- dobs %>% str_extract("(January|February|March|April|May|June|July|August|September|October|November|December)")
dobs_y <- dobs %>% str_extract("\\d{4}")
dobs <- paste(dobs_y, dobs_m, dobs_d, sep="-") %>% ymd()
dat_20w$dob <- dobs

# Calculate age at Tokyo 2020
dat_20w <- dat_20w %>% 
  mutate(age = eeptools::age_calc(dob, ymd("2021-08-07"), units = "years") %>% floor)

# Sex category 
dat_20w$sex <- "Women"
# Game label
dat_20w$olympic <- "Tokyo2020"

```

```{r, message=FALSE, warning=FALSE}
#### Men at Rio 2016 ####
# URL of Men at Rio 2016
url_16m <- "https://en.wikipedia.org/wiki/Athletics_at_the_2016_Summer_Olympics_%E2%80%93_Men%27s_marathon"

# Extract all tables in the page
tab <- read_html(url_16m) %>% 
  html_nodes("table") 
# Table of interest
dat_16m <- tab[[6]] %>% 
  html_table() 

# Data reshaping
colnames(dat_16m)[1:4] <- c("rank", "athlete", "country", "time")
dat_16m$rank[1:3] <- c(1:3)
dat_16m <- dat_16m %>% 
  mutate(rank = ifelse(rank=="—", NA, rank)) %>% 
  mutate(sb = ifelse(is.na(rank), NA, 0)) %>% 
  mutate(sb = ifelse(Notes %in% c("SB","NR","PB"), 1, sb)) %>% 
  mutate(dnf = ifelse(time=="DNF" | Notes %in% c("DNF","DNS","DSQ"), 1, 0)) %>% 
  mutate(time = ifelse(dnf==1, NA, time)) %>% 
  mutate(time = lubridate::hms(time)) %>% 
  select(rank, athlete, country, time, sb, dnf)

# Age data collection
urls <- read_html(url_16m) %>% 
  html_nodes("table") %>% 
  .[[6]] %>% 
  html_nodes("a[href *= '/wiki/' ]") %>% # extract the individual athletes' wikis
  html_attr("href") %>% 
  .[!str_detect(.,"Olympics") & !str_detect(., "Bests") & !str_detect(., "conditions") & !str_detect(., "records")] %>%
  paste("https://en.wikipedia.org", ., sep="")

# Extraction of Wiki text
wiki_text <- sapply(urls, text_detect)

# DOB detection by string processing
dobs <- sapply(wiki_text, dob_detect) %>% 
  as.character() %>% 
  str_replace("born ", "")
dobs_d <- dobs %>% str_extract("\\d{1,2}")
dobs_m <- dobs %>% str_extract("(January|February|March|April|May|June|July|August|September|October|November|December)")
dobs_y <- dobs %>% str_extract("\\d{4}")
dobs <- paste(dobs_y, dobs_m, dobs_d, sep="-") %>% ymd()
dat_16m$dob <- dobs

# Calculate age at Rio 2016
dat_16m <- dat_16m %>% 
  mutate(age = eeptools::age_calc(dob, ymd("2016-08-21"), units = "years") %>% floor)

# Sex category 
dat_16m$sex <- "Men"
# Game label
dat_16m$olympic <- "Rio2016"



```

```{r, message=FALSE, warning=FALSE}
#### Women at Rio 2016 ####
# URL of Women at Rio 2016
url_16w <- "https://en.wikipedia.org/wiki/Athletics_at_the_2016_Summer_Olympics_%E2%80%93_Women%27s_marathon"

# Extract all tables in the page
tab <- read_html(url_16w) %>% 
  html_nodes("table") 
# Table of interest
dat_16w <- tab[[6]] %>% 
  html_table() 

# Data reshaping
colnames(dat_16w)[1:4] <- c("rank", "athlete", "country", "time")
dat_16w$rank[1:3] <- c(1:3)
dat_16w <- dat_16w %>% 
  mutate(rank = ifelse(rank=="—", NA, rank)) %>% 
  mutate(sb = ifelse(is.na(rank), NA, 0)) %>% 
  mutate(sb = ifelse(Notes %in% c("SB","NR","PB"), 1, sb)) %>% 
  mutate(dnf = ifelse(time=="DNF" | Notes %in% c("DNF","DNS","DSQ"), 1, 0)) %>% 
  mutate(time = ifelse(dnf==1, NA, time)) %>% 
  mutate(time = lubridate::hms(time)) %>% 
  select(rank, athlete, country, time, sb, dnf)

# Age data collection
urls <- read_html(url_16w) %>% 
  html_nodes("table") %>% 
  .[[6]] %>% 
  html_nodes("a[href *= '/wiki/' ]") %>% # extract the individual athletes' wikis
  html_attr("href") %>% 
  .[!str_detect(.,"Olympics") & !str_detect(., "Bests") & !str_detect(., "conditions") & !str_detect(., "records")] %>%
  paste("https://en.wikipedia.org", ., sep="")

# Extraction of Wiki text
wiki_text <- sapply(urls, text_detect)

# DOB detection by string processing
dobs <- sapply(wiki_text, dob_detect) %>% 
  as.character() %>% 
  str_replace("born ", "")
dobs_d <- dobs %>% str_extract("\\d{1,2}")
dobs_m <- dobs %>% str_extract("(January|February|March|April|May|June|July|August|September|October|November|December)")
dobs_y <- dobs %>% str_extract("\\d{4}")
dobs <- paste(dobs_y, dobs_m, dobs_d, sep="-") %>% ymd()
dobs[128] <- "1980-10-12" %>% ymd()   # missing data (source: https://en.wikipedia.org/wiki/Graciete_Santana)
dat_16w$dob <- dobs

# Calculate age at Rio 2016
dat_16w <- dat_16w %>% 
  mutate(age = eeptools::age_calc(dob, ymd("2016-08-14"), units = "years") %>% floor)

# Sex category 
dat_16w$sex <- "Women"
# Game label
dat_16w$olympic <- "Rio2016"


```

```{r, message=FALSE, warning=FALSE}
#### Combine the 4 dataframes ####
dat <- rbind(dat_20m, dat_20w, dat_16m, dat_16w)

#### Rename the countries ####
dat <- dat %>% 
  mutate(country = ifelse(country=="Chinese Taipei", "Taiwan", country)) %>% 
  mutate(country = ifelse(country=="Democratic Republic of the Congo", "Congo (Kinshasa)", country))

```

```{r, message=FALSE, warning=FALSE}
#### Lockdown ####
# Source: https://en.m.wikipedia.org/wiki/COVID-19_lockdowns
non_lockdown <- c("Burundi", "Iceland", "Japan", "Nicaragua",
                  "South Korea", "Sweden", "Taiwan", "Tanzania", "Uruguay")
dat <- dat %>% 
  mutate(lockdown = ifelse(country %in% non_lockdown, 0, 1))
```

```{r, message=FALSE, warning=FALSE}
#### Prior attendance at Tokyo ####
prior_attend <- dat %>% 
  group_by(athlete) %>% 
  filter(n()>1) %>% 
  pull(athlete)
dat <- dat %>% 
  mutate(prior_attend = ifelse(olympic=="Tokyo2020" & athlete %in% prior_attend, 1, 0))

```

```{r, message=FALSE, warning=FALSE}
#### COVID-19 cases / population as of July 22, 2021 ####
# COVID-19 cases collected from [Github](https://github.com/CSSEGISandData/COVID-19).
#url <- getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
jhu <- read_csv(url) %>%
  dplyr::select(`Province/State`, `Country/Region`, `7/22/21`) %>% 
  mutate(`Country/Region2` = ifelse(is.na(`Province/State`), `Country/Region`,
                                    ifelse(`Province/State`=="Hong Kong", "Hong Kong",
                                           `Country/Region`))) %>% 
  group_by(`Country/Region2`) %>% 
  summarise(case_total = sum(`7/22/21`), .groups="drop") %>% 
  rename(country = `Country/Region2`) %>% 
  dplyr::select(country, case_total)

```

```{r, message=FALSE, warning=FALSE}
#### Population/GDP ####

# Read in Johns Hopkins UID lookup table
# Source: https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv
url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
pop <- read_csv(url)

# GDP data #
# Read in GDP data on WORLD BANK
# Source: https://api.worldbank.org/v2/en/indicator/NY.GDP.MKTP.CD?downloadformat=csv
url <- "https://api.worldbank.org/v2/en/indicator/NY.GDP.MKTP.CD?downloadformat=csv"
gdp <- read_csv("API_NY.GDP.MKTP.CD_DS2_en_csv_v2_3263806.csv")

# Join the population and the GDP data
temp <- left_join(pop %>% subset(is.na(Province_State)), 
                  gdp %>% 
                    rename(iso3 = `Country Code`,
                           gdp2016 = `2016`,
                           gdp2017 = `2017`,
                           gdp2018 = `2018`,
                           gdp2019 = `2019`,
                           gdp2020 = `2020`) %>% 
                    dplyr::select(iso3, `Country Name`, gdp2016:gdp2020),
                  by = "iso3")

jhu <- left_join(jhu, 
                 temp %>% rename(country = Country_Region), 
                 by="country")

jhu <- jhu %>% 
  mutate(case_pp = case_total/Population) %>% 
  dplyr::select(country, case_pp, gdp2016:gdp2020)

# Reclassify discrepant country names
jhu <- jhu %>% 
  mutate(country = ifelse(country %in% dat$country, country, 
                          case_when(country == "Taiwan*" ~ "Taiwan",
                                    country == "Czechia" ~ "Czech Republic",
                                    country == "Democratic Republic of the Congo" ~ "Congo",
                                    country == "United Kingdom" ~ "Great Britain",
                                    country == "Korea, South" ~ "South Korea",
                                    country == "US" ~ "United States")))

```

```{r, message=FALSE, warning=FALSE}
#### join the datasets (dat+jhu+pop+gdp) ####
# No data on cases: North Korea, Palestine, Pueruto Rico, Refugee Olympic Team
dat <- left_join(dat, jhu, by="country")

```

```{r, message=FALSE, warning=FALSE}
#### Continent ####
# URL of continent category
url <- "https://www.newworldencyclopedia.org/entry/list_of_countries_by_continent"

# Extract all tables in the page
tab <- read_html(url) %>% 
  html_nodes("table") 
# Africa
africa <- tab[[1]] %>% 
  html_table() %>% 
  .[,c(1,3)]
africa <- paste0(africa[,1], africa[,2], "South Sudan") # add discrepant country name
# Asia
asia <- tab[[2]] %>% 
  html_table() %>% 
  .[,c(1,3)]
asia <- paste0(asia[,1], asia[,2], "Taiwan") # add discrepant country name
# Europe
europe <- tab[[3]] %>% 
  html_table() %>% 
  .[,c(1,3)]
europe <- paste0(europe[,1], europe[,2], "Great Britain") # add discrepant country name
# North America
n_america <- tab[[4]] %>% 
  html_table() %>% 
  .[,c(1,3)]
n_america <- paste0(n_america[,1], n_america[,2])
# South America
s_america <- tab[[5]] %>% 
  html_table() %>% 
  .[,c(1,3)]
s_america <- paste0(s_america[,1], s_america[,2])
# Oceania
oceania <- tab[[6]] %>% 
  html_table() %>% 
  .[,c(1,3)]
oceania <- paste0(oceania[,1], oceania[,2])

dat <- dat %>% 
  mutate(continent = case_when(str_detect(africa, country) ~ "Africa",
                               country=="Congo (Kinshasa)" ~ "Africa", # add discrepant country name
                               str_detect(asia, country) ~ "Asia",
                               str_detect(europe, country) ~ "Europe",
                               str_detect(n_america, country) ~ "North America",
                               str_detect(s_america, country) ~ "South America",
                               str_detect(oceania, country) ~ "Oceania"))

# time conversion into seconds for analyses
dat <- dat %>% 
  mutate(time_sec = period_to_seconds(time)) %>% 
  select(rank:time, time_sec, sb:last_col())
```

```{r, message=FALSE, warning=FALSE}
#### Final dataset ####
rm(list = ls()[!ls()=="dat"])
dat %>% slice(1:10) %>% knitr::kable()
```

- `rank`: rank at each Olympic game
- `athlete`: athlete names
- `country`: country from which the athlete is
- `time`: records at each Olympic game
- `time_sec`: records at each Olympic game in seconds
- `sb`: season best of the athlete; including national record and personal pest
- `dnf`: did not finish; including did not start and disqualified
- `dob`: date of birth
- `age`: age (years) at the time of each Olympic game
- `sex`: sex category
- `olympic`: Olympic game name
- `case_pp`: total case per population of the athlete country as of July 22, 2021
- `gdp2016:2020`: GDP from 2016 to 2020
- `continent`: continent from which the athlete is
- `lockdown`: yes if the country from which the athletes is implemented lockdown
- `prior_attend`: prior attendance of Tokyo 2020 athletes at Rio 2016 (athletes at Rio 2016 are all "no")



## Regression Part (Satoko Ugai)##

```{r, warning=FALSE}
#This study examined whether there was a difference in marathon records between pre- and post -COVI19 Olympics. A total of 506 athletes participating Rio2016 and Tokyo2020 Olympics were included in this study.The primary outcome was a marathon record. The predictor was Rio Olympic2016 (i.e.,pre-COVID19 Olympic) and Tokyo Olympic2020 (i.e.,post-COVID19 Olympic).

#Making table1
library(gtsummary)
labels <-  list(time_sec ~ "Marathon record,mean sec (SD)", 
                sb ~ "Season best, n (%)",
                dnf ~ "Did not finish, n (%)",
                age ~ "Age, mean year(SD)",
                sex ~ "Sex,n (%)",
                continent ~ "Continent,n(%)",
                lockdown ~ "Number of athletes from locked down countries")
#making table1 overall
tbl1<- tbl_summary(data=dat %>% 
                   
                   dplyr::select(olympic,time_sec,sb,dnf,age,sex,continent,lockdown),
                   by=olympic,
                   label=labels,
                   statistic=list(c(time_sec,age)~"{mean}({sd})"),
                   missing="no") %>%
  add_p(test=list(c(time_sec,age)~"t.test",
                  all_categorical()~"chisq.test.no.correct")) %>%
                  modify_caption("**Table 1. demographic Characteristics of all athletes**") 
tbl1

#Table1 summarizes demographic characteristics of all athletes according to Rio Olympic 2016 and Tokyo Olympic 2020. Two groups were compared regarding the demographic characteristics with t test for continuous variables and chi-squared test for categorical variables. There were 312 athletes who participated in the Rio Olympics and 194 athletes who participated in the Tokyo Olympics.The mean marathon record is 9165 seconds (2H32M45sec) in Rio Olympic and 8887 seconds (2H28M07sec) in Tokyo Olympic. The t test showed that a p-value was <0.001.We therefore conclude that there is a statistically significant difference between marathon records and Olympics among male athletes.The proportion of athletes who achieved a season's best and those who dropped out of the marathon race was higher in the Tokyo Olympics than that in the Rio Olympics. By continent, the percentages of athletes from Asia and South America was lower in the Tokyo Olympics than those in the Rio Olympics.

#making table2 men
labels1 <-  list(time_sec ~ "Marathon record,mean sec (SD)", 
                sb ~ "Season best, n (%)",
                dnf ~ "Did not finish, n (%)",
                age ~ "Age, mean year(SD)",
                continent ~ "Continent,n (%)",
                lockdown ~ "Number of athletes from locked down countries")
tbl2<- tbl_summary(data=dat %>% 
                   filter(sex=="Men") %>%
                   dplyr::select(olympic,time_sec,sb,dnf,age,continent,lockdown),
                   by=olympic,
                   label=labels1,
                   statistic=list(c(time_sec,age)~"{mean}({sd})"),
                   missing="no") %>%
  add_p(test=list(c(time_sec,age)~"t.test",
                  all_categorical()~"chisq.test.no.correct")) %>%
                  modify_caption("**Table 2. demographic Characteristics of male athletes**") 
tbl2

#Table2 summarizes demographic characteristics of male athletes according to Rio Olympic 2016 and Tokyo Olympic 2020.The mean marathon record is 8542 seconds (2H22M22sec) in Rio Olympic and 8324 seconds (2H18M44sec) in Tokyo Olympic. The t test showed that a p-value was <0.001.We therefore conclude that there is a statistically significant difference between marathon records and Olympics among male athletes. The percentages of athletes who achieved season best and those who dropped out of the marathon race was significantly higher at the Tokyo Olympics than those at the Rio Olympics. Like the characteristics of overall athletes, the percentages of male athletes from Asia and South America was lower in the Tokyo Olympics than those in the Rio Olympics.

#making table3 women
tbl3<- tbl_summary(data=dat %>% 
                   filter(sex=="Women") %>%
                   dplyr::select(olympic,time_sec,sb,dnf,age,continent,lockdown),
                   by=olympic,
                   label=labels1,
                   statistic=list(c(time_sec,age)~"{mean}({sd})"),
                   missing="no") %>%
  add_p(test=list(c(time_sec,age)~"t.test",
                  all_categorical()~"chisq.test.no.correct")) %>%
                  modify_caption("**Table 3. demographic Characteristics of female athletes**") 
tbl3

#Table3 summarizes demographic characteristics of female athletes according to Rio Olympic 2016 and Tokyo Olympic 2020. The mean marathon record is 9814 seconds (2H43M34sec) in Rio Olympic and 9473 seconds (2H37M53sec)in Tokyo Olympic. The t test showed that a p-value was <0.001. We conclude that there is a statistically significant difference between marathon records and Olympics among female athletes. Similar to the male athletes, the percentage of athletes who achieved a season's best time was higher in the Tokyo Olympics than that in the Rio Olympics, while there was no statistically significant difference in the percentage of athletes who drop out of the race between the two Olympics. The percentages of athletes from Asia and South America was lower in the Tokyo Olympics than those in the Rio Olympics similarly to overall and male athletes.


#Multivariable-adjusted linear regression

#male
dat1=dat %>% 
     filter(sex=="Men")
dat$continent<-as.factor(dat$continent)

model3<-lm(time_sec ~  olympic+sb+age+continent,data=dat1)
summary(model3)
model3 %>% 
  tbl_regression(intercept = TRUE)%>%
as_gt() %>%
  gt::tab_header(title = "Table 4. Multivariable-adjusted linear regression",
                 subtitle = "Men")
#Female
dat2=dat %>% 
     filter(sex=="Women")
model4<-lm(time_sec ~  olympic+sb+age+continent,data=dat2)
summary(model4)
model4 %>% 
  tbl_regression(intercept = TRUE)%>%
as_gt() %>%
  gt::tab_header(title = "Table 5. Multivariable-adjusted linear regression",
                 subtitle = "Women")

#Making figure1
#Marathon Record in Male athlete from Asian countries

p3<-dat%>%filter(sex=="Men"|continent=="Asia") %>%
  ggplot(aes(X=factor(olympic), y=time)) +
  geom_boxplot(aes(olympic,time)) +
  scale_y_time() +
  xlab("Olympic") +
  ylab("Records" ) +
  ggtitle("Figure 1. Marathon Record in Male athlete from Asian countries")
p3


#Making figure2
#Marathon Record in Female athlete from Asian countries

p4<-dat%>%filter(sex=="Female"|continent=="Asia") %>%
  ggplot(aes(X=factor(olympic), y=time)) +
  geom_boxplot(aes(olympic,time)) +
  scale_y_time() +
  xlab("Olympic") +
  ylab("Records" ) +
  ggtitle("Figure 2. Marathon Record in Female athlete from Asian countries")
p4

#After adjusting for season-best times, age, and continent in the multivariable logistic regression model, there is no statistically significant difference in marathon times between the Rio and Tokyo Olympics for both men and women (Table 4,5). There were fewer Asian athletes participating in the Tokyo Olympics compared to the Rio Olympics. In addition, the records of Asian athletes were faster in the Tokyo Olympics than in the Rio Olympics (Figure 1,2). For example, North Korea did not participate as a country in order to protect athletes from COVID19 infection. The three women runner from North Korean who did not participate the race had marathon world rankings of 80th, 123rd, and 158th respectively. The male runner from North Korean who did not participate the race was ranked 168th. It is possible that the world marathon rankings of North Korean athletes in the Tokyo Olympics were relatively lower than the athletes who participated in the Tokyo Olympics. Therefore, it is possible that there is a selection bias for the athletes who participated in the Tokyo Olympics in this study.

#Extend the model to evaluate whether there is evidence that the association between olympic and Maratho record is different for those with and without a lockdown policy

#Male

model5<-lm(time_sec ~  olympic+sb+age+continent+olympic*lockdown,data=dat1)
summary(model5)
model5 %>% 
  tbl_regression(intercept = TRUE)%>%
as_gt() %>%
  gt::tab_header(title = "Table 5. Multivariable-adjusted linear regression (evaluate the interaction by lockdown policy)",
                 subtitle = "Men")

#Female

model6<-lm(time_sec ~  olympic+sb+age+continent+olympic*lockdown,data=dat2)
summary(model6)
model6 %>% 
  tbl_regression(intercept = TRUE)%>%
as_gt() %>%
  gt::tab_header(title = "Table 6. Multivariable-adjusted linear regression (evaluate the interaction by lockdown policy)",
                 subtitle = "Women")

#We evaluate whether there is evidence that the association between Olympics and marathon record is different between countries with and without a lockdown policy. The multivariable linear regression showed that a p-value was 0.7 for male and 0.4 for female. Therefore, we fail to reject the null hypothesis at a 0.05 level of significance and conclude that the association between marathon records and pre-/post-COVID19 Olympics does not significantly differ by the lockdown policy(Table 5,6). However, the sample size of countries without the lockdown policy is small. Therefore, it is possible that we didn't have enough power to detect a statistical interaction.

#Extend the model to evaluate whether total COVID-19 cases per population of each country is a significant predictor for Marathon records.

#Men
dat3<-dat %>% 
  filter(sex %in% "Men"|olympic%in%"Tokyo2020")
model7<-lm(time_sec ~  olympic+case_pp+sb+age+continent+olympic*lockdown,data=dat3)
summary(model7)
model7 %>% 
  tbl_regression(intercept = TRUE)%>%
as_gt() %>%
  gt::tab_header(title = "Table 7. Multivariable-adjusted linear regression",
                 subtitle = "Men in Tokyo2020")

#Women
dat4<-dat %>% 
  filter(sex %in% "Women"|olympic%in%"Tokyo2020")
model8<-lm(time_sec ~  olympic+case_pp+sb+age+continent+olympic*lockdown,data=dat4)
summary(model8)
model8 %>% 
  tbl_regression(intercept = TRUE)%>%
as_gt() %>%
  gt::tab_header(title = "Table 8. Multivariable-adjusted linear regression",
                 subtitle = "Women in Tokyo2020")

#We evaluate whether total COVID-19 cases per population of each country is a significant predictor for Marathon records.The p-values were 0.3 in male athletes and 0.006 in female athletes(Table 7,8).Therefore, we can say that the total COVID-19 cases per each country is a significant predictor for Marathon records among female athletes.It is possible that the severer the infection was, the better the marathon records were.Therefore, female athletes who participated in the Tokyo Olympics might be highly physically capable of getting better records even though they were affected by the infection.

```


```{r, warning=FALSE}

```

