---
title: "Scraping Part"
author: "Yusuke Yoshikawa (BST260 Group11)"
output: html_document
---


## Data Collection Part

### Brief sumamry of scraping
First, I collected the tables of marathon records at 4 Olympic marathon game. I further scraped the dates of birth of all athletes by accessing their individual pages. Scraping process was similar across the 4 games, so please see the descriptions at Men Tokyo2020 scraping for data collection of the Olympic games. Then, I combined the 4 datasets into one dataframe. 
Second, I collected data regarding lockdown policy, COVID-19 cases, populations, financial and geographic data of area/countries which athletes are from. 
Finally, I saved the dataset as a csv file for subsequent analyses by other members. 

```{r, message=FALSE, warning=FALSE}
#### Libraries ####
library(tidyverse)
library(rvest)
library(lubridate)
#if (!require(RCurl)) {install.packages("RCurl", dependencies=TRUE)}
#library(RCurl)
```




---
title: "Visualization Part"
author: "Marie Wu (BST260 Group11)"
output: html_document
---
## Visualization to Explore the Impact of the COVID-19 Pandemic on Olympic Marathon Performances

After we defined our variables and completed our data collection in marathon results at the Rio 2016 and Tokyo 2020 games, we further explored the basic composition of our data. In addition, we also explored how athletes' marathon finishing times were similar or different in distribution across 2016 and 2020, as well as the impact of the COVID-19 pandemic on finishing times.

The results of attendance by gender and by continent across 2016 and 2020 are shown in Graph 1 and Graph 2, respectively. The distributions of finishing times at the Rio and Tokyo games for both genders are shown in Graph 3 and Graph 4. 

The marathon results at the Rio and Tokyo games for both genders were compared by continent in Graph 5 and Graph 6. Graph 7A and 7B show the scatterplots for eyeballing men's marathon results versus GDP across 2016 and 2020; likewise, Graph 8A and 8B show the scatterplots for eyeballing women's marathon results versus GDP across 2016 and 2020. In Graphs 9 and 10, we explored the relationship between top finishing athletes and COVID-19 severity within the countries they represent via scatterplots and bar charts.

Graphs 11 and 12 allow us to view the countries from which the top 20 marathon athletes in both genders originate on COVID-19 heatmaps across 2016 and 2020.

```{r message=FALSE, warning=FALSE}
load("data.RData")
library(dplyr)
library(tidyverse)
library(gridExtra)
library(dslabs)
library(ggplot2)
library(ggthemes)
library(scales)
library(lubridate)
library(ggpubr)

#Naming df by 2016 2020 Men Female
data(dat)
dat_16m <- dat %>% filter(olympic == "Rio2016", sex=="Men")
dat_16f <- dat %>% filter(olympic == "Rio2016", sex=="Women")
dat_20m <- dat %>% filter(olympic == "Tokyo2020", sex=="Men")
dat_20f <- dat %>% filter(olympic == "Tokyo2020", sex=="Women")
dat_Male<- dat %>% filter(sex=="Men")
dat_Female<- dat %>% filter(sex=="Women")
```
